**Reference-based OTU clustering for machine learning classification**

### 2022-08-18 submission/manuscript.Rmd

${^*}$ Current Affiliation:

@@

Machine learning classification of disease based on the gut microbiome
often relies on clustering 16S rRNA gene sequences into operational
taxonomic units (OTUs) to quantify microbial composition. The standard
*de novo* approach to clustering sequences into OTUs leverages the similarity of
the sequences to each other rather than to a reference database. The
abundance of each OTU is used to train a classification model. However,
OTU assignments depend on the sequences in the data set and therefore
can change if new data are added. This lack of consistency complicates
classification because in order to use the model to classify additional
samples, the OTUs have to be reclustered to include the new sequences
and the model retrained with the new OTU clusters. A new reference-based
clustering algorithm, called OptiFit, addresses this issue by fitting
new sequences into existing OTUs. While OptiFit has been shown to produce
high quality OTU clusters, it is unknown whether this method will impact 
the performance of
classification models. We used OptiFit to cluster additional data into
existing OTU clusters and quantified model performance in classifying a
data set containing samples from patients with and without colonic
screen relevant neoplasias (SRN). We compared the performance of this
model to the standard procedure of *de novo* clustering all the data together. We
found that both approaches performed equally well in classifying SRNs.
Moving forward, when OTUs are used in classification problems, OptiFit
can streamline the process of classifying new samples by avoiding the need to 
retrain models using reclustered sequences.

@@

There is great potential for using microbiome data to non-invasively
diagnose people. One of the challenges with using classification models
generated using the relative abundance of operational taxonomic unit
(OTU) data is that 16S rRNA gene sequences are assigned to OTUs based on
their similarity to other sequences in the dataset. If data are
generated from new patients seeking a diagnosis, then it would be
necessary to reassign sequences to OTUs and retrain the classification
model. Yet there is a desire to have a single, validated model that can
be deployed. To overcome this obstacle, we applied the OptiFit
clustering algorithm which fits new sequence data to existing OTUs,
allowing the reuse of a consistent model. A random forest machine
learning model deployed using OptiFit performed as well as the
traditional reassignment and retraining approach. This result indicates
there is potential for developing machine learning classification models
based on OTU relative abundance data.

@@

Gut community composition is useful as a resource for machine learning
classification of diseases, including colorectal cancer [@baxter2016;
@zackular2014]. Amplicon sequencing of the 16S rRNA gene is a reliable
tool for assessing the taxonomic composition of microbial communities,
which is the input to these models. Analysis of 16S rRNA gene sequence
data generally relies on clustering sequences based on similarity
into operational taxonomic units (OTUs). The process of OTU clustering
can either be reference-based or *de novo*. The quality of OTUs
generated with reference-based clustering is generally poor compared to
those generated with *de novo* clustering @westcott2015. While *de novo*
clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the data in the data set and the addition of new data
could change the overall OTU clusters. The inconsistent nature of *de novo* OTU
clustering complicates deployment of machine learning models since
integration of additional data requires reclustering all the data and
retraining the model. The ability to integrate new data into a
validated model without reclustering and retraining could allow for
deployment of a single model that new data can be continually tested against.
Recently, Sovacool *et al* introduced OptiFit: a method for fitting new
sequence data into existing OTUs @sovacool2022. While OptiFit is proven
to effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on classification performance.
Here, we tested the ability of OptiFit to cluster new sequence data into
existing OTU clusters for the purpose of classification of disease based
on gut microbiome composition.

We compared two approaches, one using all of the data to generate *de novo* OTU
clusters, and the other generating *de novo* OTU clusters with a portion of the
data and then fitting the remaining sequence data to the existing OTUs
using OptiFit. In the first approach, all of the 16S rRNA sequence data
was *de novo* clustered into OTUs with the OptiClust algorithm in mothur
@westcott2017. The resulting abundance data was then split into training
and testing sets, where the training set was used to tune
hyperparameters and ultimately train the model. The testing set was then
classified with the model and the performance of the model was
quantified (Figure 1A). However, with this methodology, we would have to
regenerate the OTU clusters and retrain the model if we wanted to
classify additional samples. The OptiFit algorithm @sovacool2022
addresses this problem by enabling new sequences to be clustered into
existing OTUs. The OptiFit workflow is similar to the OptiClust workflow,
where the data was clustered into OTUs and used to tune hyperparameters
and ultimately train the model. Then, we used OptiFit to fit sequence
data of samples not part of the original data set into the existing OTUs,
and used the same model to classify the samples (Figure 1B). To test how
the model performance compares between these two methodologies, we used
a publicly available data set of 16S rRNA gene sequences from stool
samples of healthy subjects as well as subjects with SRN consisting of
advanced adenoma and carcinoma @baxter2016. The data set was randomly
split into an 80% train set and 20% test set. For the standard OptiClust
workflow, the training and test sets were *de novo* clustered together
into OTUs, then the resulting abundance table was split into the training
and testing set. For the OptiFit workflow, the train set was clustered
*de novo* into OTUs, and the remaining test set was fit to the OTU
clusters using the OptiFit algorithm. For both workflows, the abundance
table of the train set was used to tune hyperparameters and train a
random forest model to classify SRN. The test set was classified as
either control or SRN using the trained models. To account for variation
depending on the split of the data, the data set was randomly split 100
times and the process repeated for each of the 100 data splits. By
comparing the model performance of classifying the samples in the test
data set between the OptiFit and OptiClust algorithms, we quantified the
impact of using OptiFit on model classification performance.

@@

optifit_avg_mcc <- merged_mcc %>% 
  filter(algorithm == "optifit" & state=="combo") %>% 
  summarise(mean_mcc = mean(mcc))
```

We first examined the quality of the resulting OTU clusters from the two
algorithms using the Matthews correlation coefficient (MCC). The MCC
score was quantified by examining all pairs of sequences and assessing
whether they belonged together in an OTU based on their similarity
@westcott2017. MCC scores range between negative one and one. A score of
negative one means none of the sequences in an OTU are within the
similarity threshold and any sequences within the similarity threshold
are not in an OTU together. An MCC score of zero essentially means the
sequences are randomly clustered. An MCC score of 1 means all sequences
in an OTU are within the similarity threshold and all sequence pairs
within the similarity threshold are in the same OTU. To ensure that
OptiFit is appropriately integrating new sequence data into the existing
OTUs, we expected the MCC scores produced by the OptiClust and OptiFit
workflows to be similar. Since the data was only clustered once in the
OptiClust workflow there was only one MCC score while the OptiFit
workflow produced an MCC score for the OTU clusters from each data
split. Overall, the MCC scores were similar between OptiClust (MCC =
`r round(opticlust_mcc,digits=3)`) and OptiFit (average MCC =
`r round(optifit_avg_mcc,digits=3)`). This indicated that OptiFit
performed as well as OptiClust when integrating new sequences into the
existing OTUs.

@@

After verifying that the quality of the OTUs was consistent between
OptiClust and OptiFit, we examined the model performance for classifying
samples in the held out test data set. To quantify model performance, we
used the OTU relative abundances from the training data from the
OptiClust and OptiFit workflows to train a model to predict SRNs. Using
the predicted and actual diagnosis classification, we calculated the
area under the receiver operating characteristic curve (AUROC) for each
data split to quantify model performance. During cross-validation (CV)
training, the model performance was equivalent between the two
algorithms (p-value =
`r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`,
OptiClust mean CV AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_cv_AUC),3)`,
OptiFit mean CV AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_cv_AUC),3)`,
Figure 2A). The trained model was then deployed to classify the samples
of the test data as control or SRN. The performance on the test data was
equivalent between the two algorithms (p-value =
`r format_decimal(pvals %>% filter(metric == "AUC") %>% select(p_value),2)`,
OptiClust mean test AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_AUC),3)`,
OptiFit mean test AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_AUC),3)`,
Figure 2B,C) indicating that new data could be fit to existing OTU
clusters without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification using OTU
relative abundance. A potential problem with using OptiFit is that any
sequences in the new data that do not map to the existing OTU clusters
will be discarded, resulting in a possible loss of information. However,
we demonstrated that OptiFit can be used to fit new sequence data into
existing OTU clusters and performs equally well in predicting SRN
compared to *de novo* clustering all of the sequence data together. The ability to
integrate data from new samples into existing OTUs enables the
deployment of a single machine learning model. These results are based
on a single data set and disease. Further analysis is needed to
determine the number of samples that are necessary to build a robust
model capable of classifying diverse samples. A robust machine learning
model could be implemented as part of a non-invasive and low-cost aid in
diagnosing SRN and other diseases.

@@

***Data Processing.*** The full dataset was preprocessed with mothur
(v1.47) @schloss2009 to join forward and reverse reads, merge duplicate
reads, align to the SILVA reference database (v132) @quast2013,
precluster, remove chimeras with UCHIME @edgar2011, assign taxonomy, and
remove non-bacterial reads following the Schloss Lab MiSeq standard
operating procedure described on the mothur website
(<https://mothur.org/wiki/miseq_sop/>). 100 splits of the 490 samples
were generated where 80% of the samples (392 samples) were randomly
assigned to the training set and the remaining 20% (98 samples) were
assigned to the test set. Using 100 splits of the data accounts for the
variation that may be observed depending on the samples that are in the
training or test sets. Each sample was in the training set an average of
`r n_test_train %>% pull(avg_train)` times
(SD=`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and
the test set an average of `r n_test_train %>% pull(avg_test)` times
(SD=`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).

The data was processed through two workflows. First, the standard
workflow using the OptiClust algorithm @westcott2017. In this workflow,
all of the data was clustered together with OptiClust to generate OTUs
and the resulting abundance tables were split into the training and
testing sets. In the second workflow, the preprocessed data was split
into the training and testing sets. The training set was clustered into
OTUs, then the test set was fit to the OTUs of the training set using
the OptiFit algorithm @sovacool2022. The OptiFit algorithm was run with the open
method so that any sequences that did not map to the existing OTU
clusters would form new OTUs. 
<!-- Did you then remove the columns corresponding to the additional OTUs? 
Typically you'd want to run the closed method for this use-case, right? -->
For both workflows, the shared files were
sub-sampled to 10,000 reads per sample.

***Machine Learning.*** Machine learning using Random Forest was
conducted with the R package mikrompl (v 1.2.0) @topçuoglu2021 to
predict the diagnosis (SRN or normal) for the samples in the test set
for each data split. The training set was preprocessed to normalize OTU
counts (scale and center), collapse correlated OTUs, and remove OTUs with
zero variance. The preprocessing from the training set was then applied
to the test set. Any OTUs in the test set that were not in the training
set were removed. P values comparing model performance were calculated
as previously described @topçuoglu2020. The averaged ROC curves were
plotted by taking the average and standard deviation of the sensitivity
at each specificity value.

@@

**Figure 1: Workflows.** **A)** OptiClust workflow: The full data set
was clustered into OTUs using the OptiClust algorithm in mothur. The
data was then split into two sets where 80% of the samples were assigned
to the training set and 20% to the testing set. The training set was
preprocessed with mikropml to normalize values (scale and center), collapse
correlated features, and remove features with zero variance. Using
mikropml, the training set was split into train and validate sets to
compare results using different hyperparameter settings. The highest
performing hyperparameter setting was then used to train the model with
the full training set. The preprocessing scale from the training set was
applied to the test data set, then the trained model was used to
classify the samples in the test set. Based on the actual classification
and predicted classification, the area under the receiver operating
characteristic curve (AUROC) was calculated to summarize model
performance. The entire process was repeated 100 times to account for
variability depending on the split of the data resulting in a total of
100 AUROC values summarizing the performance of the standard OptiClust
workflow. **B)** OptiFit workflow: The data set was first split into two
sets where 80% of the samples were assigned to the training set and 20%
to the testing set. The training set was then clustered into OTUs using
the OptiClust algorithm in mothur. The resulting abundance data was
preprocessed with mikropml to normalize values (scale/center), collapse
correlated features, and remove features with zero variance. Using
mikropml, the training set was split into train and validate sets to
compare results using different hyperparameter settings. The highest
performing hyperparameter setting was then used to train the model with
the full training set. The OptiFit algorithm in mothur was used to
cluster the held out testing data set using the OTUs of the training set
as a reference. The preprocessing scale from the training set was
applied to the test data set, then the trained model was used to
classify the samples in the test set. Based on the actual classification
and predicted classification, the area under the receiver operating
characteristic curve (AUROC) was calculated to summarize model
performance. The entire process was repeated 100 times to account for
variability depending on the split of the data resulting in a total of
100 AUROC values summarizing the performance of the new OptiFit
workflow.

@@

**Figure 2: Model Performance.** **A)** Area under the receiver
operating characteristic (AUROC) curve during cross-validation for the
OptiClust and OptiFit workflows. Mean and standard deviation of the
AUROC is represented by the black dot and whiskers. Mean AUROC is
printed to the right of the points. **B)** AUROC on the test data for
the OptiClust and OptiFit workflows. Mean and standard deviation of the
AUROC is represented by the black dot and whiskers. Mean AUROC is
printed to the right of the points. **C)** Receiver operating
characteristic (ROC)Averaged ROC curves

>>>>

${^*}$ Current Affiliation: Bio-Rad Laboratories, Hercules, California,
USA

@@

Machine learning classification of disease based on the gut microbiome
often relies on clustering 16S rRNA gene sequences into operational
taxonomic units (OTUs) to quantify microbial composition. The abundance
of each OTU is then used to train a classification model. The standard
*de novo* approach to clustering sequences into OTUs leverages the
similarity of the sequences to each other rather than to a reference
database. However, OTU assignments depend on the sequences in the data
set and therefore can change if new data are added. This lack of
stability complicates classification because in order to use the model
to classify additional samples, the new sequences must be reclustered
with the old data and the model must be retrained with the new OTU
assignments. The new reference-based clustering algorithm, OptiFit,
addresses this issue by fitting new sequences into existing OTUs. While
OptiFit can produce high quality OTU clusters, it is unclear whether
this method for fitting new sequence data into existing OTUs will impact
the performance of classification models. We used OptiFit to cluster
additional data into existing OTU clusters and then quantified model
performance in classifying a data set containing samples from patients
with and without colonic screen relevant neoplasia (SRN). We compared
the performance of this model to the standard procedure of *de novo*
clustering all the data together. We found that both approaches
performed equally well in classifying SRNs. Moving forward, when OTUs
are used in classification problems, OptiFit can streamline the process
of classifying new samples by avoiding the need to retrain models using
reclustered sequences.

@@

There is great potential for using microbiome data to non-invasively
diagnose people. One of the challenges with using classification models
based on the relative abundance of operational taxonomic units (OTUs) is
that 16S rRNA gene sequences are assigned to OTUs based on their
similarity to other sequences in the data set. If data are generated
from new patients seeking a diagnosis, then it would be necessary to
reassign sequences to OTUs and retrain the classification model. Yet
there is a desire to have a single, validated model that can be widely
deployed. To overcome this obstacle, we applied the OptiFit clustering
algorithm which fits new sequence data to existing OTUs and allows for
the reuse of a consistent model. A random forest machine learning model
deployed using OptiFit performed as well as the traditional reassignment
and retrain approach. This result indicates that there is potential for
developing deployable machine learning classification models based on
OTU relative abundance data.

@@

Gut community composition is useful as a resource for machine learning
classification of diseases, such as colorectal cancer [@baxter2016;
@zackular2014]. Taxonomic composition of microbial communities can be
assessed using amplicon sequencing of the 16S rRNA gene, which is the
input to classification models. Analysis of 16S rRNA gene sequence data
generally relies on similarity-based clustering of sequences into
operational taxonomic units (OTUs). The process of OTU clustering can
either be reference-based or *de novo*. The quality of OTUs generated
with reference-based clustering is generally poor compared to those
generated with *de novo* clustering @westcott2015. While *de novo*
clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the sequences within the data set and the addition of
new data has the potential to redefine OTU cluster composition. The
unstable nature of *de novo* OTU clustering complicates deployment of
machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for deployment of a single model that new data
can be continually tested against. Recently, Sovacool *et al* introduced
OptiFit, a method for fitting new sequence data into existing OTUs
@sovacool2022. While OptiFit can effectively fit new sequence data to
existing OTU clusters, it is unknown if the use of OptiFit will have an
impact on classification performance. Here, we tested the ability of
OptiFit to cluster new sequence data into existing OTU clusters for the
purpose of classifying disease based on gut microbiome composition.

We compared two approaches, one using all the data to generate OTU
clusters, and the other generating *de novo* OTU clusters with a portion
of the data and then fitting the remaining sequence data to the existing
OTUs using OptiFit. In the first approach, all the 16S rRNA sequence
data was *de novo* clustered into OTUs with the OptiClust algorithm in
mothur @westcott2017. The resulting abundance data was then split into
training and testing sets, where the training set was used to tune
hyperparameters and ultimately train the model. The testing set was then
classified with the model and the performance of the model was
quantified (Figure 1A). However, with this methodology, we would have to
regenerate the OTU clusters and retrain the model if we wanted to
classify additional samples. The OptiFit algorithm @sovacool2022
addresses this problem by enabling new sequences to be clustered into
existing OTUs. The OptiFit workflow is similar to the OptiClust
workflow, where the data was clustered into OTUs and used to tune
hyperparameters and ultimately train the model. Then, we used OptiFit to
fit sequence data of samples not part of the original data set into the
existing OTUs, and used the same model to classify the samples (Figure
1B). To test how the model performance compared between these two
approaches, we used a publicly available data set of 16S rRNA gene
sequences from stool samples of healthy subjects (n = 261) as well as
subjects with SRN consisting of advanced adenoma and carcinoma (n = 229)
@baxter2016. The data set was randomly split into an 80% train set and
20% test set. For the standard OptiClust workflow, the training and test
sets were *de novo* clustered together into OTUs, then the resulting
abundance table was split into the training and testing set. For the
OptiFit workflow, the train set was clustered *de novo* into OTUs, and
the remaining test set was fit to the OTU clusters using the OptiFit
algorithm. For both workflows, the abundance table of the train set was
used to tune hyperparameters and train a random forest model to classify
SRN. The test set was classified as either control or SRN using the
trained models. To account for variation, the data set was randomly
split 100 times and the process repeated for each of the 100 data
splits. By comparing the model performance of classifying the samples in
the test data set between the OptiFit and OptiClust algorithms, we
quantified the impact of using OptiFit on model classification
performance.

@@

optifit_avg_mcc <- merged_mcc %>% 
  filter(algorithm == "optifit" & state=="combo") %>% 
  summarise(mean_mcc = mean(mcc),
            sd_mcc = sd(mcc))
```

We first examined the quality of the resulting OTU clusters from the two
algorithms using the Matthews correlation coefficient (MCC). The MCC
score was quantified by examining all pairs of sequences and assessing
whether they belonged together in an OTU based on their similarity
@westcott2017. MCC scores range between negative one and one. A score of
negative one means none of the sequences in an OTU are within the
similarity threshold and any sequences within the similarity threshold
are not in an OTU together. An MCC score of zero means the sequences are
randomly clustered. An MCC score of 1 means all sequences in an OTU are
within the similarity threshold and all sequence pairs within the
similarity threshold are in the same OTU. To ensure that OptiFit is
appropriately integrating new sequence data into the existing OTUs, we
expected the MCC scores produced by the OptiClust and OptiFit workflows
to be similar. Since the data was only clustered once in the OptiClust
workflow there was only one MCC score while the OptiFit workflow
produced an MCC score for the OTU clusters from each data split.
Overall, the MCC scores were similar between OptiClust (MCC =
`r round(opticlust_mcc,digits=3)`) and OptiFit (average MCC =
`r round(optifit_avg_mcc %>% pull(mean_mcc),digits=3)`, standard
deviation = `r round(optifit_avg_mcc %>% pull(sd_mcc),digits=3)`). This
indicated that OptiFit performed as well as OptiClust when integrating
new sequences into the existing OTUs.

@@

After verifying that the quality of the OTUs was consistent between
OptiClust and OptiFit, we examined the model performance for classifying
samples in the held out test data set. To quantify model performance, we
used the OTU relative abundances from the training data from the
OptiClust and OptiFit workflows to train a model to predict SRNs. Using
the predicted and actual diagnosis classification, we calculated the
area under the receiver operating characteristic curve (AUROC) for each
data split to quantify model performance. During cross-validation (CV)
training, the model performance was equivalent between the two
algorithms (p-value =
`r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`,
OptiClust mean CV AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_cv_AUC),3)`,
OptiFit mean CV AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_cv_AUC),3)`;
Figure 2A). The trained model was then deployed to classify the samples
of the test data as control or SRN. The performance on the test data was
equivalent between the two algorithms (p-value =
`r format_decimal(pvals %>% filter(metric == "AUC") %>% select(p_value),2)`,
OptiClust mean test AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_AUC),3)`,
OptiFit mean test AUROC =
`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_AUC),3)`;
Figures 2B and 2C) indicating that new data could be fit to existing OTU
clusters without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification using OTU
relative abundance. A potential problem with using OptiFit is that any
sequences from the new samples that do not map to the existing OTU
clusters will be discarded, resulting in a possible loss of information.
Regardless, we demonstrated that OptiFit can be used to fit new sequence
data into existing OTU clusters and performs equally well in predicting
SRN compared to *de novo* clustering all the sequence data together. The
ability to integrate data from new samples into existing OTUs enables
the deployment of a single machine learning model. Further analysis is
needed to determine the number of samples that are necessary to build a
robust model capable of classifying diverse samples. A robust machine
learning model could be implemented as part of a non-invasive and
low-cost aid in diagnosing SRN and other diseases.

@@

***Data Processing.*** The full data set was preprocessed with mothur
(v1.47) @schloss2009 to join forward and reverse reads, merge duplicate
reads, align to the SILVA reference database (v132) @quast2013,
precluster, remove chimeras with UCHIME @edgar2011, assign taxonomy, and
remove non-bacterial reads following the Schloss Lab MiSeq standard
operating procedure described on the mothur website
(<https://mothur.org/wiki/miseq_sop/>). 100 splits of the 490 samples
were generated where 80% of the samples (392 samples) were randomly
assigned to the training set and the remaining 20% (98 samples) were
assigned to the test set. Using 100 splits of the data accounts for the
variation that may be observed depending on the samples that are in the
training or test sets. Each sample was in the training set an average of
`r n_test_train %>% pull(avg_train)` times (standard deviation =
`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and the
test set an average of `r n_test_train %>% pull(avg_test)` times
(standard deviation =
`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).

The data was processed through two workflows. In the control workflow,
all the data was clustered together with OptiClust @westcott2017 to
generate OTUs and the resulting abundance tables were split into the
training and testing sets. In the experimental workflow, the
preprocessed data was split into the training and testing sets. The
training set was clustered into OTUs using OptiClust, then the test set
was fit to the OTUs of the training set using the OptiFit algorithm
@sovacool2022. The OptiFit algorithm was run with method open so that
any sequences that did not map to the existing OTU clusters would form
new OTUs. For both pathways, the shared files were sub-sampled to 10,000
reads per sample.

***Machine Learning.*** Machine learning using random forest was
conducted with the R package mikrompl (v 1.2.0) @topçuoglu2021 to
predict the diagnosis (SRN or normal) for the samples in the test set
for each data split. The training set was preprocessed to normalize OTU
counts (scale and center), collapse correlated OTUs, and remove OTUs
with zero variance. The preprocessing from the training set was then
applied to the test set. Any OTUs in the test set that were not in the
training set were removed. P values comparing model performance were
calculated as previously described @topçuoglu2020. The averaged ROC
curves were plotted by taking the average and standard deviation of the
sensitivity at each specificity value.

@@

**Figure 1: Workflows.** **A)** OptiClust workflow: The full data set
was clustered into OTUs using the OptiClust algorithm in mothur. The
data was then split into two sets where 80% of the samples were assigned
to the training set and 20% to the testing set. The training set was
preprocessed with mikropml to normalize values (scale and center),
collapse correlated features, and remove features with zero variance.
Using mikropml, the training set was split into train and validate sets
to compare results using different hyperparameter settings. The highest
performing hyperparameter setting was then used to train the model with
the full training set. The preprocessing scale from the training set was
applied to the test data set, then the trained model was used to
classify the samples in the test set. Based on the actual classification
and predicted classification, the area under the receiver operating
characteristic curve (AUROC) was calculated to summarize model
performance. The entire process was repeated 100 times to account for
variability depending on the split of the data resulting in a total of
100 AUROC values summarizing the performance of the standard OptiClust
workflow. **B)** OptiFit workflow: The data set was first split into two
sets where 80% of the samples were assigned to the training set and 20%
to the testing set. The training set was then clustered into OTUs using
the OptiClust algorithm in mothur. The resulting abundance data was
preprocessed with mikropml to normalize values (scale/center), collapse
correlated features, and remove features with zero-variance. Using
mikropml, the training set was split into train and validate sets to
compare results using different hyperparameter settings. The highest
performing hyperparameter setting was then used to train the model with
the full training set. The OptiFit algorithm in mothur was used to
cluster the held out testing data set using the OTUs of the training set
as a reference. The preprocessing scale from the training set was
applied to the test data set, then the trained model was used to
classify the samples in the test set. Based on the actual classification
and predicted classification, the area under the receiver operating
characteristic curve (AUROC) was calculated to summarize model
performance. The entire process was repeated 100 times to account for
variability depending on the split of the data resulting in a total of
100 AUROC values summarizing the performance of the new OptiFit
workflow.

@@

**Figure 2: Model Performance.** **A)** Area under the receiver
operating characteristic (AUROC) curve during cross-validation for the
OptiClust and OptiFit workflows. Mean and standard deviation of the
AUROC is represented by the black dot and whiskers. Mean AUROC is
printed to the right of the points. **B)** AUROC on the test data for
the OptiClust and OptiFit workflows. Mean and standard deviation of the
AUROC is represented by the black dot and whiskers. The mean AUROC is
printed to the right of the points. **C)** Averaged receiver operating
characteristic (ROC) curves.
